Why Did Russia Fare So Poorly during COVID: An Analysis of Why One of the Best Prepared Countries Could Not Control Excess Deaths

### 2024-02-14 paper/paper.qmd
New Sections and Feedback Draft
- Added new section based on peer review

---
title: "Why Did Russia Fare So Poorly during COVID: An Analysis of Why One of the Best Prepared Countries Could Not Control Excess Deaths"
author: 
  - Aryaman Sharma
  - Aviral Bhardwaj
  - Janel Gilani
thanks: "Code and data are available at: https://github.com/Ary4m3n/covid-effect-russia.git; Replication on Social Science Reproduction platform available at: https://www.socialsciencereproduction.org/reproductions/8938b287-6681-4a5e-8993-42011a51be58/index"
date: today
date-format: long
abstract: " Despite Russia being deemed well-prepared for pandemics, it faced an unprecedented rise in death rates compared to other nations. This study aims to replicate Nuzzo and Ledesma's findings, examining COVID-19's impact on Russia while considering variables such as the Global Health Security (GHS) Index and excess death rates. Our analysis revealed a significantly elevated excess death rate in Russia, contrasting with the lower rates in more prepared countries. This research sheds light on the COVID-19 outbreak's impact on Russia and identifies areas for improvement for well-prepared countries to mitigate future pandemics' effects, emphasizing the importance of better planning and regulations."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

As mentioned above, the data analysed in this paper is from the reproduction package of the paper by Nuzzo and Ledesma [@citeRepPaper]. The data was cleaned and analysed using the open source `R` programming language [@citeR]. R libraries and packages such as `tidyverse` [@citeTidyverse], `janitor` [@citeJanitor], `ggplot2` [@citeGgplot], `knitr` [@citeKnitr], `readr` [@citeReadr], `dplyr` [@citeDplyr], `rnaturalearth` [@citeRnaturalearth] and `sf` [@citeSf]. Since, we have replicated the process of data cleaning and analysis, the data used in this paper is the same as the data used in the original paper.

@@

In addition to tracking deaths, a crucial data point during the COVID-19 pandemic is the excess death rate, representing the number of deaths occurring within a specific period above the expected baseline. This metric serves as a key indicator of the pandemic's impact on mortality. In our dataset, death rates are standardized to a per 100,000 population basis, facilitating meaningful comparisons across countries.

The process of collecting this data is primarily overseen by institutions like the Institute for Health Metrics and Evaluation (IHME), which collates information from various sources available for each country. Deaths included in the excess death rate calculation encompass those directly attributed to COVID-19, as well as those indirectly influenced by the pandemic where COVID-19 appeared on the death certificate. For instance, deaths resulting from delayed treatment due to healthcare system strain caused by the pandemic are also accounted for.

@@

Once observed deaths are verified and categorized, they are tallied according to established criteria for inclusion as defined by IHME. the excess death rate calculation. This meticulous approach ensures the accuracy and reliability of the data used to compute the excess death rate for each country.

However, due to data truncation, the actual death rate isn't always precise, leading to inaccuracies in calculating the excess death rate. This introduces measurement errors, undermining the accuracy of both the actual death rate and the excess death rate. This is a significant challenge in accurately measuring the pandemic's impact on a country.

<!-- When discussing COVID-19, (1) issues with underreported or missing data, particularly related to non-responses, are prevalent, (2) the lack of a standardized definition of COVID-19 deaths, which can lead to measurement errors (3) vastness of countries poses logistical challenges in reporting and conducting surveys, making it difficult to gather comprehensive data (4) variations in testing methodologies and availability of resources across regions can further distort the data. -->

<!-- Many countries, especially developing nations like India, Brazil, and Russia, poses above issue. This poses a significant challenge in accurately measuring the pandemic's impact on a country. For example, the excess death rate, a variable in our analysis, measures the number of deaths occurring in a specific period above the expected number. This metric helps estimate the pandemic's impact. However, due to data truncation, the actual death rate isn't always precise, leading to inaccuracies in calculating the excess death rate. This introduces measurement errors, undermining the accuracy of both the actual death rate and the excess death rate. -->

<!-- With respect to measurement and accounting for the GHS index, it encompasses various measurements grouped into six broad categories: (1) prevention, (2) detection and reporting, (3) rapid response, (4) health system, (5) compliance with international norms, and (6) risk environment. However, various measurement errors and qualitative biases can arise from the subjective nature of these categories. For example, the "prevention" category includes measurements such as "vaccination coverage" and "biosecurity and biosafety." However, the definition of "biosecurity and biosafety" can vary across countries, leading to measurement errors. Additionally, the GHS index is based on self-reported data, which can be subject to biases and inaccuracies. This can lead to measurement errors in the GHS index, which can further distort the analysis. -->

@@

This section will be divided into three parts where all together we aim to look at Russia's capabilty according to the GHS index to tackle pandemics (@sec-results-ghs), then compare Russia's trend in all cause death rates and excess death rates with those of other countries like India, South Korea and the United states (@sec-results-death-rates) and finally study the relationship between the excess death rate and GHS index for Russia along with a subset of countries in the world (@sec-results-death-rates).

@@

This observation leads us to @fig-excess-death-rate, where we present the excess death rate for the four countries for 2020 and 2021. As mentioned above, the excess death rates were calculated by extrapolating the dashed line of best-fit from years 2010-2019 in @fig-all-cause-death-rate representing the trend in all cause death rates to 2020 and 2021. This way we were able to find the expected death rate for each of the four countries and subsequently we were able to plot the difference between the expected death rate and the actual death rate called the excess death rate for 2020 and 2021 for all four countries in @fig-excess-death-rate.

@@

## Relationship between the Excess Death Rate and Global Health Security Index (GHS) {#sec-results-death-rates}

>>>>

---
title: "Why Did Russia Fare So Poorly during COVID: An Analysis of Why One of the Best Prepared Countries Could Not Control Excess Deaths"
author: 
  - Aryaman Sharma
  - Aviral Bhardwaj
  - Janel Gilani
thanks: "Code and data are available at: https://github.com/Ary4m3n/covid-effect-russia.git; Replication on Social Science Reproduction platform available at: https://www.socialsciencereproduction.org/reproductions/8938b287-6681-4a5e-8993-42011a51be58/index"
date: today
date-format: long
abstract: "Despite Russia being deemed well-prepared for pandemics, it faced an unprecedented rise in death rates compared to other nations. This study aims to replicate Nuzzo and Ledesma's findings, examining COVID-19's impact on Russia while considering variables such as the Global Health Security (GHS) Index and excess death rates. Our analysis revealed a significantly elevated excess death rate in Russia, contrasting with the lower rates in more prepared countries. This research sheds light on the COVID-19 outbreak's impact on Russia and identifies areas for improvement for well-prepared countries to mitigate future pandemics' effects, emphasizing the importance of better planning and regulations."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

As mentioned above, the data analysed in this paper is from the reproduction package of the paper by Nuzzo and Ledesma [@citeRepPaper]. The data was cleaned and analysed using the open source `R` programming language [@citeR]. R libraries and packages such as `tidyverse` [@citeTidyverse], `janitor` [@citeJanitor], `ggplot2` [@citeGgplot], `knitr` [@citeKnitr], `readr` [@citeReadr], `dplyr` [@citeDplyr], `rnaturalearth` [@citeRnaturalearth] and `sf` [@citeSf]. Since, we have replicated the process of data cleaning and analysis, the data used in this paper is the same as the data used in the original paper. The estimands for this analysis are preparedness of countries to tackle pandemic, specifically Russia, and their degree of success in controlling the pandemic. To estimate these, we utilize metrics such as the Global Health Security (GHS) Index and the excess death rate respectively.

@@

A crucial data point for the COVID-19 pandemic is the excess death rate, representing the number of deaths occurring within a specific period above the expected baseline. The process of collecting this data is primarily overseen by Institute for Health Metrics and Evaluation (IHME), which collates information from various sources available for each country. Deaths included in the excess death rate calculation consists of those directly attributed to COVID-19, or in which COVID-19 appeared on the death certificates. For instance, deaths resulting from delayed treatment due to healthcare system strain caused by the pandemic are also accounted for.

@@

However, discussions surrounding COVID-19 present several challenges, including: (1) issues related to underreported or missing data, which can occur due to societal stigma at the local level, or countries intentionally concealing data to portray preparedness; (2) the absence of a universally accepted definition for COVID-19 deaths, leading to potential measurement inaccuracies; (3) the vast geographical expanse of countries, posing logistical hurdles in data collection and reporting efforts; and (4) disparities in testing methodologies and resource availability across regions, further complicating data interpretation.

Many countries, particularly nations such as India, Brazil, and Russia, which had the highest death count, faced the aforementioned challenges. This presents a significant obstacle in accurately gauging the pandemic's toll within a country. For instance, the excess death rate, a key metric in our analysis, serves to estimate the pandemic's impact. However, due to data truncation and measurement errors, the actual death rate may not always be precisely captured. Consequently, inaccuracies arise in calculating the excess death rate, introducing measurement errors that undermine the reliability of both the actual death rate and the excess death rate in data analysis.

@@

The results section is divided into four parts where all together we aim to look at Russia's capability according to the GHS index to tackle pandemics (@sec-results-ghs), then compare Russia's death rates trends with those of other countries like India, South Korea and the United states (@sec-results-death-rates), then using those trends we dive into Russia's excess death rates while comparing them with other countries (@sec-results-excess-rates) and finally study the relationship between the excess death rate and GHS index for Russia along with a subset of countries in the world (@sec-results-relationship).

@@

## Excess Death Rates {#sec-results-excess-rates}
This observation leads us to @fig-excess-death-rate, where we present the excess death rate for the four countries for 2020 and 2021. As mentioned above, the excess death rates were calculated by extrapolating the dashed line of best-fit from years 2010-2019 in @fig-all-cause-death-rate representing the trend in all cause death rates to 2020 and 2021. This way we were able to find the expected death rate for each of the four countries and subsequently we were able to plot the difference between the expected death rate and the actual death rate called the excess death rate for 2020 and 2021 for all four countries in @fig-excess-death-rate.

@@

## Relationship between the Excess Death Rate and Global Health Security Index (GHS) {#sec-results-relationship}
